---
title: "data_analysis"
output: html_document
date: "2025-04-15"
---

Descriptive statistics for my nominal and ordinal variables, which would be Family_History, Doctor_said_diabetes, What_A1C_should_be, Glyco_cat, and BMIcat. Starting by just running some frequency tables for all.
```{r}
project_data <- read.csv("/Users/avahoffer/Documents/anth_241_final/clean_data")
```

```{r}
library(dplyr)
count(project_data, Family_History)
count(project_data, Doctor_said_diabetes)
count(project_data, What_A1C_should_be)
count(project_data, Glyco_cat)
count(project_data, BMIcat)
```
wow! now we're gonna do some descriptive stats for my continuous variables, which are BMI, Glyco_Percent, Age, and Moderate. We're gonna get some means and medians through our Summary function :)
```{r}
#object_name <- c("variable1", "variable2", etc.)
#summary(data[objectname])
contin_variables <- c("BMI", "Glyco_Percent", "Age", "Moderate")
summary(project_data[contin_variables])
```
now for some more descriptive stats for our continuous variables.
```{r}
library(tidyr)
library(dplyr)
#data%>% 
#drop_na(variable)%>% 
#summarize(mean(Variable), sd(Variable), count=n())
project_data%>%
  drop_na(BMI)%>%
  summarize(mean(BMI), sd(BMI), count=n())

project_data%>%
  drop_na(Glyco_Percent)%>%
  summarize(mean(Glyco_Percent), sd(Glyco_Percent), count=n())

project_data%>%
  drop_na(Age)%>%
  summarize(mean(Age), sd(Age), count=n())

project_data%>%
  drop_na(Moderate)%>%
  summarise(mean(Moderate), sd(Moderate), count=n())
```


Now to actually display our nominal and ordinal variables, we're gonna do some BAAARRR CHAAARRRTTTTS!
```{r}
library(ggplot2)
#a. Bar chart for a nominal variable showing raw counts: Create the bar chart for BMI categories (BMIcat)
#ggplot (data, aes(x=Catvar)) + geom_bar() + labs(title="", x="", y="")

ggplot(project_data, aes(x=new_fam_history)) + geom_bar(fill="pink") + labs("Bar Chart of Diabetes Family History", x="Family History", y="Count")

ggplot(project_data, aes(x=Doctor_said_diabetes)) + geom_bar(fill="pink") + labs("Bar Chart Table of Doctor Saying Diabetes", x="If Participants Doctor Said They Have Diabetes", y="Count")

ggplot(project_data, aes(x=Glyco_cat)) + geom_bar(fill="pink") + labs("Bar Chart of Categorized A1C Levels", x="Categorized A1C Level", y="Count")

ggplot(project_data, aes(x=What_A1C_should_be))+ geom_bar(fill="pink")+ labs("Bar Chart of What Doctor Said A1C Levels Should Be", x="A1C Level", y="Count")

ggplot(project_data, aes(x=BMIcat))+geom_bar(fill="pink")+ labs("Bar Chart of BMI Categories", x="BMI Category", y="Count")

```
Right well now we have displayed some of our data, but what if we compared some nominal and ordinal variables? Time to make a grouped bar chart! Specifically, a group bar chart of Glyco_cat and BMIcat.

```{r}

#grouped bar chart:
#ggplot(data, aes(x=Dependent variable, fill=Independent variable)) + geom_bar(position = "dodge") + labs(title="", x="Gender", y="Frequency")

ggplot(project_data, aes(x=Glyco_cat, fill=BMIcat))+ geom_bar(position="dodge") + labs(title= "Bar Chart of Glyco Category and BMI Category", x="Comparison", y="Frequency")

```
Wow! It looks like people with diabetes and prediabetes have more obese people compared to the other BMI categories, but is that significant? Thats a question for later!

Now lets compare Family_History and Glyco_cat
```{r}
library(ggplot2)
#grouped bar chart:
#ggplot(data, aes(x=Dependent variable, fill=Independent variable)) + geom_bar(position = "dodge") + labs(title="", x="Gender", y="Frequency")

ggplot(project_data, aes(x=Glyco_cat, fill= new_fam_history))+ geom_bar(position="dodge") + labs(title="Bar Chart Comparing Family History and Glycohemoglobin Category", x= "Comparison", y="Frequency")

```
Huh, well thats interesting. It seems that some people with diabetes dont really have a family history of diabetes. 

But now its time to see some variation in our continuous variables!
Lets do a box and whisker plot!

```{r}
#a.ggplot(data, aes(x="", y=Contvar)) + geom_boxplot() + labs(y="", x="", title="")

ggplot(project_data, aes(y=Glyco_Percent))+ geom_boxplot()+labs(y="Glycohemogoblin Percentage in Blood", title = "Box Plot of Glycohemogoblin")
```
Well that wasnt very helpful because its so itty bitty, maybe a histogram?
Then I might do some more for the rest of my continuous variables, as that is a requirement.
```{r}
#colored histogram with 20 bins: ggplot(data, aes(x=contvar)) + geom_histogram(bins=20, color="white", fill="steelblue") + labs(title="Colored histogram with 20 bins", x="", y="")

ggplot(project_data, aes(x=Glyco_Percent))+ geom_histogram(bins = 40, color="white", fill="pink")+ labs(title = "Glycohemogoblin Percent Histogram", x="Percentage", y="Count")

ggplot(project_data, aes(x=BMI))+ geom_histogram(bins = 40, color="white", fill="pink")+ labs(title = "BMI Histogram", x="BMI", y="Count")

ggplot(project_data, aes(x=Age))+ geom_histogram(bins = 40, color="white", fill="pink")+ labs(title = "Age Histogram", x="Age", y="Count")

ggplot(project_data, aes(x=Moderate))+ geom_histogram(bins = 40, color="white", fill="pink")+ labs(title = "Moderate Minutes of Exercise Histogram", x="Minutes Per Day", y="Count")


```

Now its time to compare some continuous by categorical/nominal variables >:)
T tests on Glyco_Percent by Family_History!!!
```{r}
#just gonna get some means and sds here
project_data%>%
  drop_na(Glyco_Percent)%>%
  group_by(new_fam_history)%>%
  summarize(mean(Glyco_Percent), sd(Glyco_Percent), count=n())

hist(project_data$Glyco_Percent)
#var.test(Contvar~Catvar, data=, alternative="two.sided")

t.test(Glyco_Percent~new_fam_history, var.equal=FALSE, data=project_data)
```

Having issues with that :( lets just move on for now
NEVERMIND WE LOVE MALENA IT WORKED!!! AND ITS SIGNIFIIICANNNNNNNT!
KRUSKAL WALACE TIME!
```{r}
project_data%>%
  drop_na(Glyco_Percent)%>%
  group_by(BMIcat)%>%
  summarize(mean(Glyco_Percent), sd(Glyco_Percent), count=n())

#dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)

dunn.test(x=project_data$Glyco_Percent, g=project_data$BMIcat, kw=TRUE)
```
Wow, so basically everything is significant, and the overall p value is 0. 

Thats cool, but what I feel like will be more revealing is a test with BMI and Glyco_cat, to see if BMI has an effect on for glyco level and whether or not you have diabetes.

```{r}
project_data%>%
  drop_na(BMI)%>%
  group_by(Glyco_cat)%>%
  summarize(mean(BMI), sd(BMI), count=n())
#dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)

dunn.test(x=project_data$BMI, g=project_data$Glyco_cat, kw=TRUE)

```
Wow! So there are significant differences between all the categories. 

Heres where im gonna try my what_doctor_said dunn/kruskal test

```{r}
project_data%>%
  drop_na(Glyco_Percent)%>%
  group_by(Doctor_said_diabetes)%>%
  summarize(mean(Glyco_Percent), sd(Glyco_Percent), count=n())

dunn.test(x=project_data$Glyco_Percent, g=project_data$Doctor_said_diabetes, kw=TRUE)
```


Time to compare some nominal and ordinal significane levels with aaaaaaaa... CHI SQUARE TEST!!
I think doing Glyco_cat by Family_history would be good, if Family_History works that is

```{r}
#a. Generate a two-way frequency table of BMIcat and HHFoodSec:
#table_name <- xtabs(~ CatVar1 + CatVar2, data=)
#print the table
#show the table in percentages: prop.table(table_name)*100
Glycocat_FamilyHistory <- xtabs(~Glyco_cat + new_fam_history, data=project_data)
print(Glycocat_FamilyHistory)
prop.table(Glycocat_FamilyHistory)*100
#b. run the chi-square test:
#chisq.test(table_name)
chisq.test(Glycocat_FamilyHistory)
```
Family_History was being mean again, so we're gonna try a Chi square test with something else. 
NOW IT WORKS! Wowza, and tis significant, which I guess it should be because glyco percent and fam history was significant too

```{r}
Glycocat_BMIcat <- xtabs(~Glyco_cat+ BMIcat, data=project_data)
print(Glycocat_BMIcat)
prop.table(Glycocat_BMIcat)*100

chisq.test(Glycocat_BMIcat)
```
Okay, so, the p-value is waaaaaay less than 0.05, so there is a relationship between glyco cat and bmi cat. 


I would also like to do a test to see if the difference between what the doctor said and glyco cat is significant. I want to know if people are aware that they have diabetes.

```{r}
#a. Generate a two-way frequency table of BMIcat and HHFoodSec:
#table_name <- xtabs(~ CatVar1 + CatVar2, data=)
#print the table
#show the table in percentages: prop.table(table_name)*100

Glycocat_Doctorsaid <- xtabs(~Glyco_cat + Doctor_said_diabetes, data=project_data)
print(Glycocat_Doctorsaid)

#b. run the chi-square test:
#chisq.test(table_name)

chisq.test(Glycocat_Doctorsaid)
```
So the p value is significant, so there is a relationship between glyco cat and what the doctor said. Maybe ANOVA would be good for this? Have to wait for Malena. Actually, I might go up and try a dunn test with it.

Now, time to do some correlationnns! I think first I'll test to see if theres a correlation between age and glyco percent
```{r}
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()

ggplot(aes(x=Glyco_Percent, y=BMI), data=project_data)+
  geom_point(color="pink")+
  geom_smooth(method=lm, color="blue")+
  labs(x="Glycohemogoblin Percent", y="Age")+
  theme_bw()

#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")

cor.test(project_data$Glyco_Percent, project_data$BMI, method="pearson")

```
Okay, so the correlation coefficient is 0.18, and the p value is waaaaaaaaay low. 
Now to try the same thing with glyco percent and moderate minutes of activity

```{r}
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()

ggplot(aes(x=Glyco_Percent, y=Moderate), data=project_data)+
  geom_point(color="yellow")+
  geom_smooth(method=lm, color="lightpink")+
  labs(x="Glycohemogoblin Percent", y="Minutes of Moderate Exercise Per Day")+
  theme_bw()

#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")

cor.test(project_data$Glyco_Percent, project_data$Moderate, method="pearson")
```
Well that one was not significant because the p-value is 0.21, which is greater than 0.05. 

Now we're gonna use some linear regression models to see if we can use a continuous independent variable to test a dependent continuous variable!

```{r}
#model_name <- lm(DepVar~IndVar, data=)
#summary(model_name)

Moderate_GlycoPercent <- lm(Glyco_Percent~Moderate, data=project_data)
summary(Moderate_GlycoPercent)
```
Okay, so the regression equation is Glyco_Percent= 5.78 + Moderate(-0.0005)
The r-squared value is 0.0007, and the p-value is 0.21.
From these results we know that there is no significant relationship between minutes of moderate actvity and glycohemogoblin percentage, and you have around a 0% chance of predicting someones Glyco percent based off of how much exercise they do. 

Now I wanna see if you can predict someones Glyco percent based off of their BMI

```{r}
#model_name <- lm(DepVar~IndVar, data=)
#summary(model_name)

Glycopercent_BMI <- lm(Glyco_Percent~BMI, data=project_data)
summary(Glycopercent_BMI)
```
Regression equation: Glyco_Percent= 5 + BMI(0.03)
The r-squared is 0.03, and the p-value is way less than 0.05.
So there is a significant relationship, but you have around a 0% chance of predicting someones Glyco percent based off of their BMI. 

Now I want to calculate some odds ratios, to see the odds of having diabetes based on family history and bmi. But in order to do this, I have to ask for help. 

Glyco_Percent is a continuous variable that I then converted to ordinal variable, Glyco_cat. I want to convert glyco_cat to a "Yes" or "No" question.

```{r}
#1. Create the odds table:
#odds_table_name <- table(data$IndVar, data$DepVar)

odds_BMI_Glcoyn <- table(project_data$BMIcat, project_data$new_glyco_yn)

#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)

oddsratio(odds_BMI_Glcoyn)

#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")

ggplot(data=project_data)+
  geom_mosaic(aes(x=product(BMIcat), fill=new_glyco_yn), na.rm=TRUE, show.legend = FALSE)+
  labs(title="Odds of Having Diabetes Based on BMI Category", x="BMI Category", y="Diabetes")
```
If you are obese you have a 2.31 times chance of developing diabetes compared to healthy people. If you are overweight you have a 1.72 times chance of developing diabetes. If you are underweight you have a 0.01 times chance of developing diabetes. 

The chance of developing diabetes while obese is significant because the p value is 0.000000000000036 which is significant, the chance of developing diabetes while overweight is 0.00000053 which is significant, and the chance of developing diabetes while underweight is significant because the p value is 0.00012. 

Now I would like to try it by family history again, to see if it works.



```{r}
#1. Create the odds table:
#odds_table_name <- table(data$IndVar, data$DepVar)

oddstable_glycoyn_by_doctorsaid <- table(project_data$Doctor_said_diabetes, project_data$new_glyco_yn)

#2. Run the oddsratio on the table you created:
#oddsratio(odds_table_name)

oddsratio(oddstable_glycoyn_by_doctorsaid)

#3. Graph the results in a mosaic table:
#ggplot(data=) + geom_mosaic(aes(x=product(IndVar), fill=DepVar), na.rm=TRUE, show.legend=FALSE) + labs(title="", x="", y="")

ggplot(data=project_data)+
  geom_mosaic(aes(x=product(Doctor_said_diabetes), fill=new_glyco_yn), na.rm=TRUE, show.legend = FALSE)+
  labs(title="Odds Table of Having Diabetes Based on What Your Doctor Said", x="What the Doctor Said", y="Diabetes")
```
I might try an ANOVA now, to see if its easier to understand than the dunn test.

```{r}
#1. find summary statistics of sample sizes, means, standard deviations, and 95% confidence intervals
project_data_reduced<- project_data %>% 
  drop_na(Glyco_cat,BMI)

plotdata<- project_data_reduced%>%
  group_by(Glyco_cat)%>%
  summarize(n=n(),
            mean = mean(BMI),
            sd = sd(BMI),
            ci=qt(0.975, df=n-1)*sd/sqrt(n))
plotdata

#2. test for group differences (ANOVA)
#fit <- aov(Age~BMIcat, data=NHANESdata_reduced)

fit <- aov(BMI~Glyco_cat, data=project_data_reduc)
summary(fit)

#3. Plot group means and confidence intervals to visualize these group differences.
#library(ggplot2)

ggplot(plotdata, aes(x=Glyco_cat, y=mean, group=1)) + 
  geom_point(size=1, color="pink") +
  geom_errorbar(aes(ymin=mean-ci,
                    ymax=mean+ci),
                width=.1)+
  theme_bw()+
  labs(x="Glycohemoglobin Category",
       y="BMI",
       title="Glycohemoglobin by BMI")

```
So it is significant, but im not sure why my graph isnt working. But since its significant, we'll do the turkey tukey thing.

```{r}
#1. calculates pairwise comparisons as a Tukey post hoc test
pairwise <- TukeyHSD(fit)
pairwise
#2. create a data set of the above results
BMI_Glycocat_plot <- as.data.frame(pairwise[[1]])
BMI_Glycocat_plot$conditions <- row.names(BMI_Glycocat_plot)
#3. Plot the results
ggplot(data=BMI_Glycocat_plot, aes(x=conditions, y=diff)) +
  geom_point(size=2, color="pink") +
  geom_errorbar(aes(ymin=lwr, ymax=upr, width=.2)) +
  geom_hline(yintercept=0, color="pink", linetype="dashed") +
  labs(y="Differences in mean levels", x="",
       title="95% family-wise confidence levels", x="Glyco Category", y="BMI") +
  theme_bw() +
  coord_flip()
```

